---
title: "To Bayes or to Bootstrap - that is the question"
author: "Ullrika Sahlin and Dmytro Perepolkin, Lund University"
date: "18 October 2019"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

```

# About this tutorial 

This tutorial compares quantifying epistemic uncertainty by Bayesian inference with the Bootstrap method. Bootstrap is a method that frequentists often use to get estimation errors or probability intervals to their estimates, when that doesn’t follow automatically from the analysis. The Bootstrap method generates a sample from a test statistic/parameter or variable of interest. You always get a probability distribution or sample with Bayesian inference. We discuss in what way it may be easier or more difficult to embrace a bootstrap compared to a Bayesian analysis using a general mixed model for illustration. Our conclusion is that Bayesian inference is a coherent principle for inference and more straightforward to implement than the Bootstrap method added to a frequentist analysis. 

The tutorial produced for **BayesDays at Liverpool 2019**.
```{r ll, message=FALSE, error=FALSE, warning=FALSE}
library(readr)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr) # note that you need to update to current CRAN version
library(modelr)
library(tidybayes)
library(lme4)
library(brms)
library(future)
theme_set(theme_minimal())

```
## Read in data
We will use a data set on birds in agricultural landscape. The data is taken from this publication: 
*Stjernman, M., Sahlin, U., Olsson, O., and Smith, H. G.. 2019. Estimating effects of arable land use intensity on farmland birds using joint species modeling. Ecological Applications 29( 4):e01875. 10.1002/eap.1875*

### Load data

Let's load the data and have a look at it.
```{r bird}
df <- read_csv2("data/DataEcologicalApplications_jan2019.csv")

glimpse(df)
```
## Model description
We want to build a model that predict **abundance of skylark** as a function of **agricultural area** and **landscape complexity**. The effect of agricultural land and landscape complexity on skylark is treated as a fixed effect. response to regression scale.

Data are collected in different sites (**landscapes**) and by different birdwatchers (**observer**). Unexplained variation due to landscape and observer are treated as random effects. 

Since the response is a count variable and we have both fixed and random effects we will use a Generalized Linear Mixed Model (GLMM).

Before continuing, we will scale one of the predictors (to improve convergence), give a generic name on the response variable (to easily replace it with other species) and create a categorical variable of one of the predictions (for plotting).
```{r addbird}
df <- df %>% 
  mutate(sFarml_area = as.vector(scale(Farml.area))) %>% 
  rename(birdcount = countSk) %>% 
  mutate(landscape_type = cut(LHI, breaks=c(-10,0.5,10),
                              labels=c('simple','complex')))
```

### Plot the data
```{r studydata, fig.width=10}
p1 <- ggplot(df, aes(x = Farml.area, y = birdcount)) + 
  geom_point(size = 3,aes(colour = landscape_type))+
  geom_smooth(method = loess)+
  theme(legend.position="bottom")

p2 <- ggplot(df, aes(x = LHI, y = birdcount)) + 
  geom_point(size = 3,aes(colour = Farml.area))+
  geom_smooth(method = loess)+
  theme(legend.position="bottom")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```

# Classical inference 
Let us first see how this model can be parametrised using classical (likelihood-based) inference. We estimate parameters of the GLMM using `glmer()` in the `lme4` R-package. 
```{r mod}
mod <- glmer(birdcount ~ sFarml_area * LHI + (1|landscape) + (1|observer) + offset(log(npoint)), 
             data = df, family = poisson)
summary(mod)
```

## Extracting information from the model
We can extract the fixed effects using *fixef(mod)* and random effects using *ranef(mod)*. In order to get the variance for the random effects we write 
```{r var_randomeffect}
getME(mod,'theta')
```

## Check the model 
### Study residuals and goodnes-of-fit
```{r res, fig.width=10}
par(mfrow=c(1,2)) 
hist(resid(mod)); plot(resid(mod))

p1 <- ggplot(df, aes(x=birdcount, y=resid(mod))) + 
  geom_point(size=3, aes(colour = landscape_type))+
  geom_smooth(method=loess) +
  geom_hline(aes(yintercept = 0),col='red')+
  theme(legend.position = "bottom")

df <- df %>% 
  add_predictions(mod, var="fit", type='response')

p2 <- ggplot(df,aes(x=fit,y=birdcount)) +
         geom_point(size = 3, aes(colour = landscape_type)) +
         geom_abline(col='blue',lwd = 1.12)+
  theme(legend.position = "bottom")
gridExtra::grid.arrange(p1, p2, ncol=2)
```

## Making predictions
### Predict training data
Make predictions of bird counts for the data we have. Predictions can be made with estimated random effects, but are provided without errors or uncertainty ranges.  

What data or information do we have in the different columns?
```{r pred_of_data}
df %>% 
  select(birdcount) %>% 
  mutate(all_random_effects=predict(mod, type='response', re.form=NULL),
         landscape_random_effect=predict(mod,type = 'response',re.form = ~(1|landscape)),
         no_random_effects=predict(mod,type = 'response',re.form = NA))

```

### Predict new data
Make predictions of bird counts for new data. First we create a new data set. Then we use the predict-function. The random effect is zero by default and still no uncertainty provided with the predictions.
```{r pred_of_newdata}
newdata <- data_grid(df, sFarml_area = seq_range(sFarml_area, 20),
                         LHI = c(-2, 1), npoint = 16) %>% 
    mutate(landscape_type=cut(LHI,breaks=c(-10,0.5,10),
                            labels=c('simple','complex')))

birdcount <- predict(mod, newdata=newdata, re.form = NA, type = 'response')
head(birdcount)
```

## Use bootstrap to predict with uncertainty 
We can perform model-based (semi-)parametric bootstrap of our mixed model (Poisson regression) using using `bootMer()` in `lme4` package.

### Predict with estimation error
```{r boot}
bm <- bootMer(mod, FUN = function(.){predict(., newdata=newdata, re.form = NA, type = 'response')}, nsim = 100 )
```

The bootstrap returns a sample for each prediction. 
```{r, fig.width=10}
head(bm$t[,1:6])
hist(bm$t[,1])
```

Plot bird abundance over farmland area.
```{r predplot,fig.width=10}
df_pred1 <- newdata %>%
  add_draws(draws = bm$t, value = ".prediction")

df %>% 
  ggplot(aes(x = sFarml_area, y = birdcount, color = landscape_type, fill = landscape_type)) +
  geom_point(aes(x = sFarml_area, y = birdcount)) +
  stat_lineribbon(data=df_pred1, aes(y = .prediction), .width = c(.95, .80, .50), alpha = 1/4) +
  ylim(c(0,125)) +
  facet_wrap(~landscape_type) +
  labs(title = 'Bird abundance over farmland area', subtitle = "Frequentist inference + Bootstrap")
```

### Predict with estimation error and landscape variability 
To include uncertainty from landscape random effect we draw random normally distributed values to the predictions. 
```{r boot_r}
bm.withlandscaperandom <- bootMer(mod, FUN = function(.){
  log_count <- predict(., newdata=newdata, re.form = NA, type = 'link') + rnorm(nrow(newdata),0,sqrt(getME(mod,'theta')[1]))
  exp(log_count)}, nsim = 100 )
```

### Plot
```{r predplot2, fig.width=10}
df_pred2 <- newdata %>% 
  add_draws(draws = bm.withlandscaperandom$t, value = ".prediction")

df %>% 
  ggplot(aes(x = sFarml_area, y = birdcount, color = landscape_type, fill = landscape_type)) +
  geom_point(aes(x = sFarml_area, y = birdcount)) +
  stat_lineribbon(data=df_pred2, aes(y = .prediction), .width = c(.95, .80, .50), alpha = 1/4) +
  ylim(c(0,125)) +
  facet_wrap(~landscape_type) +
  labs(title = 'Bird abundance over farmland area including landscape variability', subtitle = "Frequentist inference + Bootstrap")

```

# Bayesian inference
A Bayesian version of the GLMM is implemented in the `brms` R-package, which uses almost the same syntax as `lme4`. A stan model is created as a C++ code and the brm-function calls STAN for **MCMC sampling**. 

```{r bayes}
future::plan("multiprocess")
bmod <- brm(birdcount ~ sFarml_area * LHI + (1|landscape) + (1|observer) + offset(log(npoint)), 
            data = df, family = poisson(), future = TRUE, silent = TRUE)
```

## Study the model
```{r bayes_post}
summary(bmod)
```

## Make predictions with uncertainty
```{r bayes_pred}
df_pred_b1 <- df %>%
  data_grid(sFarml_area = seq_range(sFarml_area, n = 20), 
            LHI = c(-2,1), landscape = NA, observer = NA, npoint = 16) %>%
  mutate(landscape_type=cut(LHI,breaks=c(-10,0.5,10), labels=c('simple','complex'))) %>% 
    add_predicted_draws(bmod, value =".prediction", re_formula = NA)

```

### Plot 
```{r bayes_plot, fig.width=10}
df %>%
  ggplot(aes(x = sFarml_area, y = birdcount, color = landscape_type, fill = landscape_type)) +
  geom_point(aes(x = sFarml_area, y = birdcount)) +
  stat_lineribbon(data=df_pred_b1, aes(y = .prediction), .width = c(.95, .80, .50), alpha = 1/4) +
  ylim(c(0,125)) +
  facet_wrap(~landscape_type) + 
  labs(title = 'Bird abundance over farmland area', subtitle = "Bayesian inference")
```

## Predict with uncertainty including landscape variability 
```{r bayes_pred2}
df_pred_b2 <- df %>%
  data_grid(sFarml_area = seq_range(sFarml_area, n = 20), 
            LHI = c(-2,1), landscape = NA, observer = NA, npoint = 16) %>%
  mutate(landscape_type=cut(LHI,breaks=c(-10,0.5,10), labels=c('simple','complex'))) %>% 
  add_predicted_draws(bmod, value =".prediction", re_formula = ~(1|landscape),allow_new_levels = TRUE) 
```

### Plot 
```{r bayes_plot2, fig.width=10}
df %>%
  ggplot(aes(x = sFarml_area, y = birdcount, color = landscape_type, fill = landscape_type)) +
  geom_point(aes(x = sFarml_area, y = birdcount)) +
  stat_lineribbon(data=df_pred_b2, aes(y = .prediction), .width = c(.95, .80, .50), alpha = 1/4) +
  ylim(c(0,125)) +
  facet_wrap(~landscape_type) + 
  labs(title = 'Bird abundance over farmland area including landscape variability', subtitle = "Bayesian inference")

```

# Comparison
Is there a difference in predictions using the two approaches?
 

